---
title: "Vignette for FLAMES on Windows"
author: "Oliver Voogd"
package: FLAMES                
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Vignette for FLAMES on Windows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Windows FLAMES Pipeline
Windows is capable of running the FLAMES pipeline however, as the genome alignment in FLAMES is handled using minimap2, these functions are not able to be run on a Windows OS. This vignette provides a workflow for running the package primarily on Windows, with the required non-Windows functions being handled by a supported system - either MacOS or a Linux distrubution.

In order to run the FLAMES pipeline on Windows, ensure that the arguments do_genome_align and do_read_realign are set to FALSE. This will run the FLAMES pipeline, without the initial read alignment using Minimap2 step, and the later read realignment using Minimap2 step.
```{r eval=FALSE}
bulk_long_pipeline(..., do_genome_align=FALSE, do_read_realign=FALSE)
# OR
sc_long_pipeline(..., do_genome_align=FALSE, do_read_realign=FALSE)
```

Alternatively, the FLAMES pipeline can be manually executed using exported functions from the FLAMES package, and minimap2 alignment and realignment can be performed externally to complete the full pipeline.

This process for manual execution with external alignment is outlined below.

## Manual execution of FLAMES pipeline


### Environment setup
Begin by storing the required variables `annot`, `fastq`, `outdir`, `genome_fa`, in the current workspace. Ensure `outdir` exists, and is writable. More information on the values of these variables can be obtained by executing `?FLAMES::bulk_long_pipeline` or `?FLAMES::sc_long_pipeline`. The optional `config_file` variable can also be stored, or the default configuration file, located in the `extdata/` folder in the base level of this packaged can be used.

This vignette uses `BiocFileCache` to download the required variables from an example data set.
Below is the process for creating the BiocFileCache and downloading the required files. The code below is included for demonstration and is not required in order to run the Flames pipeline on Windows. This pipeline runs the bulk version of the FLAMES pipeline, however code is provided to run the single cell pipeline instead. Running the single cell pipeline
requires little modification to this process beyond the initial pipeline setup.
The downloaded files are stored in a temporary directory.
```{r eval=TRUE, echo=TRUE}
# download required files using BiocFileCache
temp_path <- tempfile()
bfc <- BiocFileCache::BiocFileCache(temp_path, ask=FALSE)
file_url <- 
  "https://raw.githubusercontent.com/OliverVoogd/FLAMESData/master/data"
annot <- bfc[[names(BiocFileCache::bfcadd(bfc, "Annotation", 
                                          paste(file_url, 
                                                "SIRV_isoforms_multi-fasta-annotation_C_170612a.gtf", 
                                                sep="/")))]] # [[ notation is used to get the local file path of the downloaded file
genome_fa <- bfc[[names(BiocFileCache::bfcadd(bfc, 
                                              "Genomefa", 
                                              paste(file_url, 
                                                    "SIRV_isoforms_multi-fasta_170612a.fasta", 
                                                    sep="/")))]]

# download the two fastq files, move them to a folder to be merged together
fastq1 <- bfc[[names(BiocFileCache::bfcadd(bfc, "Fastq1", paste(file_url, "fastq/sample1.fastq.gz", sep="/")))]]
fastq2 <- bfc[[names(BiocFileCache::bfcadd(bfc, "Fastq2", paste(file_url, "fastq/sample2.fastq.gz", sep="/")))]]
fastq_dir <- paste(temp_path, "fastq_dir", sep="/") # the downloaded fastq files need to be in a directory to be merged together
dir.create(fastq_dir)
file.copy(c(fastq1, fastq2), fastq_dir)
unlink(c(fastq1, fastq2)) # the original files can be deleted

# setup other environment variables
config_file <- system.file("extdata/SIRV_config_default.json", package="FLAMES") # the configuration file is included with the FLAMES package
outdir <- tempfile() # create a temporary output directory
if (!dir.exists(outdir)) {
    dir.create(outdir)
}
```

```{r eval=TRUE, echo=FALSE, results='hide'}
# download files generated using minimap2 (these files need to be generated by the user on a system with access to Minimap2. More information is in the section [Genome Alignment](#genome-alignment-using-minimap2))
# files are renamed so that FLAMES can find related .bam.bai files using the same file name. BiocFileCache automatically gives the files a random prefix which affects this.
genome_bam <- system.file("extdata/align2genome.bam", package="FLAMES")

realign_bam <- paste0(temp_path, "/realign2transcript.bam")
file.rename(bfc[[names(BiocFileCache::bfcadd(bfc, "Realign BAM", paste(file_url, "realign2transcript.bam", sep="/")))]], realign_bam)

realign_index <- paste0(temp_path, "/realign2transcript.bam.bai")
file.rename(bfc[[names(BiocFileCache::bfcadd(bfc, "Realign BAM Index", paste(file_url, "realign2transcript.bam.bai", sep="/")))]], realign_index)
```



### FLAMES Execution


#### Pipeline Setup
The FLAMES windows pipeline is broken into three steps, and three functions: `bulk_windows_pipeline_setup` or `sc_windows_pipeline_setup`, to setup
the environment variables for the pipeline, and those specific for the bulk or single cell pipeline, `windows_pipeline_isoforms`, which
is run after read alignment, in order to identify isoforms, and finally `windows_pipeline_quantification`, run after read realignment
to quantify the results and output a SummarizedExperiemnt or SingleCellExperiment object.

`bulk_windows_pipeline_setup` or `sc_windows_pipeline_setup` **must** be run before the other steps are begun.
```{r eval=TRUE}
library(FLAMES)
# If running the FLAMES single cell pipeline:
# pipeline_variables <- sc_windows_pipeline_setup(annot=annot, fastq=fastq_dir, outdir=outdir,
#                                                 genome_fa=genome_fa, config_file=config_file)
# or, if running the bulk pipeline:
pipeline_variables <- bulk_windows_pipeline_setup(annot=annot, fastq=fastq_dir, outdir=outdir, 
                                                  genome_fa=genome_fa, config_file=config_file)

```
The return list `pipeline_variables` contains all the required variables to run the future pipeline steps.
This initial function sets up the pipeline, and validates input. After it is run, minimap2 alignment is required using 
the files given in `pipeline_variables$return_files`. This should be undertaken on a system with access to minimap2.
After this is performed, the BAM file (and index file) returned from minimap2 should be stored with the same path
as given by `pipeline_variables$return_files$output_bam`, with the index file having the same name with additional extension `.bai`.
Once this is completed, the pipeline can resume.

#### Find Isoforms
After read alignment, the pipeline can resume by calling `windows_pipeline_isoforms`, and passing in the `pipeline_variables` list
returned from the previous step.
```{r eval=TRUE, echo=FALSE, results='hide'}
pipeline_variables$genome_bam = genome_bam
```
```{r eval=TRUE}
pipeline_variables <- windows_pipeline_isoforms(pipeline_variables)
```
Following this step, read realignment needs to be undertaken as before, on a system with access to minimap2. The resulting `.bam` and `.bam.bai` files
need to be located at `pipeline_variables$return_files$out_bam`, so the the pipeline can proceed.


#### Transcript Quantification
Finally, transcript quantification can be performed, as required:
```{r eval=TRUE, echo=FALSE, results='hide'}
pipeline_variables$realign_bam = realign_bam
```
```{r eval=TRUE}
se <- windows_pipeline_quantification(pipeline_variables)
```
The directory `outdir` now contains several output files returned from this pipeline. The output files generated by this pipeline are:

\itemize{
  \item{transcript_count.csv.gz}{ - a transcript count matrix (also contained in the SummarizedExperiment)}
  \item{isoform_annotated.filtered.gff3}{ - isoforms in gff3 format (also contained in the SummarizedExperiment)}
  \item{transcript_assembly.fa}{ - transcript sequence from the isoforms}
  \item{align2genome.bam}{ - sorted BAM file with reads aligned to genome}
  \item{realign2transcript.bam}{ - sorted realigned BAM file using the transcript_assembly.fa as reference}
  \item{tss_tes.bedgraph}{ - TSS TES enrichment for all reads (for QC)}
 }

The pipeline also returns a SummarizedExperiment  or SingleCellExperiment object, depending on the pipeline run, containing the data from `transcript_count.csv.gz`and `isoform_annotated.filtered.gff3`.
```{r eval=TRUE}
se
```

#### Session Info
```{r echo=FALSE}
utils::sessionInfo()
```